---
title: "An Example of Topic Modeling"
author: "Giulio Beltramin"
date: "2026-10-09"
output: pdf_document
---
# Topic Modelling with Multinomial - Dirichlet Model

In this notebook, we see an example of Topic Modelling using the Multinomial - Dirichlet model. We recall that this model is defined as follows:

Let $X_i$ be the word $i$ in the document we are analyzing, for $i = 1, \dots, n$. The law for $X_i$ is the following:
$$
    \mathbb{P}(X_i = j) = \lambda_j \qquad j = 1, \dots, \lvert V \rvert,
$$
where $\lvert V \rvert$ is the size of the vocabulary and $j$ is a label for the word.

Of course, in a document words will have repetitions. Since the order in which a word appear is not important, we can define $Y_j$ being the frequency of word $j$ in the document.

## Bayesian Model

\begin{align*}
    \left( y_1, \dots, y_{n} \right) \mid \boldsymbol{\lambda} &\stackrel{\tiny{iid}}{\sim} \operatorname{Multinomial}\left(\boldsymbol{\lambda}\right); \\
    \boldsymbol{\lambda} &\sim \operatorname{Dirichlet}\left(\boldsymbol{\alpha}\right).
\end{align*}

We recall that:

\begin{align*}
    f\left(\boldsymbol{y} \mid \boldsymbol{\lambda}\right) &= \frac{n!}{y_1! \dots y_{\lvert V \rvert}!} \prod_{j=1}^{\lvert V \rvert} \lambda_{j}^{y_j}, &\qquad\pi\left(\boldsymbol{\lambda}\right) &= \frac{1}{B\left(\boldsymbol{\alpha}\right)} \prod_{j = 1}^{\lvert V \rvert} \lambda_{j}^{\alpha_j - 1},
\end{align*}

with $y_j \in \mathbb{N} \cup \{0\}$ for all $j$, $\sum_{j = 1}^{\lvert V \rvert} y_j = n$, $\alpha_j \geq 0$ for all $j$ and $\sum_{j=1}^{\lvert V \rvert}\alpha_j = 1$.

```{r, include=TRUE}
# Load required libraries
library(httr2)
library(jsonlite)
library(tidytext)
library(stopwords)
library(dplyr)
library(ggplot2)

# Query Wikipedia API for Roberto Baggio's page extract
url <- "https://en.wikipedia.org/w/api.php"

req <- request(url) %>%
  req_headers("User-Agent" = "Mozilla/5.0") %>%
  req_url_query(
    action = "query",
    format = "json",
    titles = "Roberto_Baggio",
    prop = "extracts",
    explaintext = "1"
  )

response <- req_perform(req)
data <- resp_body_json(response)

# Extract raw text from the response structure
page_id <- names(data$query$pages)[[1]]
text <- data$query$pages[[page_id]]$extract
```
```{r, include = TRUE}
# turn the text into token and keep only the words, eliminating punctuation
tokens <- unlist(strsplit(tolower(text), "[^a-z]+"))
tokens <- tokens[tokens != ""]

# eliminate english stopword: words commonly used that are not useful to assess the meaning of a text, like "the", "a"...
stopw <- stopwords::stopwords("en")

tokens <- tokens[!tokens %in% stopw]
```

Let us print the 10 most probable words...

```{r, include = TRUE}
# Count frequencies and get top 10 most common words
counts <- sort(table(tokens), decreasing = TRUE)
head(counts, 10)
```

... and **always** remember to plot your data!

```{r, include = TRUE}
df_counts <- as.data.frame(counts)
colnames(df_counts) <- c("word", "count")

ggplot(df_counts, aes(x = count)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Word Frequency Distribution",
    x = "Frequency (Count)",
    y = "Number of Words"
  )
```

```{r, include = TRUE}
# Let's keep the words with ferqeuncy between 6 and 99 
counts_filtered <- counts[counts > 5 & counts < 100]

df_filtered <- data.frame(count = as.numeric(counts_filtered))

ggplot(df_filtered, aes(x = count)) +
  geom_histogram(bins = 90, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    x = "Frequency (Count)",
    y = "Number of Words"
  )
```


## Bayesian analysis

### Posterior Distribution

We saw in the previous TA that the posterior distribution is:

$$ \pi\left(\boldsymbol{\lambda} \mid \boldsymbol{y} \right) \sim \operatorname{Dirichlet}\left(\boldsymbol{\alpha} + \boldsymbol{y}\right)$$

```{r, include = TRUE}
# Extract vocabulary (token names) and calculate its length
vocabulary <- names(counts_filtered)
len_v <- length(vocabulary)

# Frequency counts array
y <- as.numeric(counts_filtered)

# Symmetric Dirichlet prior
alpha <- rep(1, len_v)

```


### Predictive Distribution

$$ p(x_{n+1} = j \mid x_1, \ldots, x_n) = \frac{\alpha_j + y_j}{n + \sum_{h = 1}^{|V|} \alpha_h}$$

for every $j = 1, \dots, \lvert V \rvert$

```{r, include = TRUE}
# Vectorized implementation
pred_dist <- (alpha + y) / (sum(alpha) + sum(y))

```

```{r, include = TRUE}
# Get the 5 most probable words
ind <- order(pred_dist, decreasing = TRUE)[1:5]
for (i in ind) {
  cat(sprintf("word: %s, prob: %s\n", vocabulary[i], pred_dist[i]))
}
```

```{r, include = TRUE}
# Get the 5 least probable words
ind <- order(pred_dist, decreasing = FALSE)[1:5]
for (i in ind) {
  cat(sprintf("word: %s, prob: %s\n", vocabulary[i], pred_dist[i]))
}
```

### Test randomness

$$
 H_0: \lambda = \left(\frac{1}{|V|} \cdots \frac{1}{|V|}\right) \quad vs. \quad H_0: \lambda \neq \left(\frac{1}{|V|} \cdots \frac{1}{|V|}\right)
$$

Let $\boldsymbol{\lambda_0} = \left(\frac{1}{|V|} \cdots \frac{1}{|V|}\right)$, then

$$
BF_{01} = \frac{L(\boldsymbol{y} | \boldsymbol{\lambda_0})}{\int_{S^{|V|}} L(\boldsymbol{y} | \boldsymbol{\lambda}) \pi(\boldsymbol{\lambda}) \mathrm{d}\boldsymbol{\lambda}} = \frac{L(\boldsymbol{y} | \boldsymbol{\lambda_0})}{m(\boldsymbol{y})}
$$

We saw in the previous lab that $m(\boldsymbol{y}) = \frac{n!}{y_1! \dots y_{\lvert V \rvert}!}\cdot\frac{B(\boldsymbol{\alpha + y})}{B(\boldsymbol{\alpha})}$. We can therefore see that
$$
BF_{01} = \frac{\prod_{j = 1}^{|V|}\left(\frac{1}{|V|}\right)^{y_j}}{\frac{B(\boldsymbol{\alpha + y})}{B(\boldsymbol{\alpha})}}.
$$
```{r, include = TRUE}
multi_beta_fn <- function(alpha) {
  prod(gamma(alpha)) / gamma(sum(alpha))
}

lambda0 <- rep(1 / len_v, len_v)

num <- prod(lambda0 ^ y)
den <- multi_beta_fn(alpha + y) / multi_beta_fn(alpha)

print(log10(num / den))
```

\texttt{R} returns a NaN (not a number). This (almost) always happens when we try to compute products of terms involving factorials.

*What to do?* **Move computations to log scale!** In this case:

```{r, include = TRUE}
# Define log10 gamma function
log10gamma <- function(x) {
  lgamma(x) / log(10.0)
}

# Define log10 multivariate beta function
log_multi_beta_fn <- function(alpha) {
  sum(log10gamma(alpha)) - log10gamma(sum(alpha))
}

log_num <- sum(y * log10(lambda0))
log_den <- log_multi_beta_fn(alpha + y) - log_multi_beta_fn(alpha)

print(log_num - log_den)
```
