---
title: "Pseudo Random Number Generators"
author: "Giulio Beltramin"
date: "2026-10-09"
output: html_document
---
# Sampling from the Uniform distribution

## Linear Congruential Generator (1958, Thomson & Rotenberg)

The Linear Congruential Generator is of of the first Pseudo-Random Number Generators (PRNGs).
It involves only very simple instructions.

Let $m$ (modulus), $\mu$ (multiplier), $c$ (increment) and $X_0$ (seed) integers.

The LCG iterates

$$
X_{i+1} = (\mu X_{i} + c) \% m
$$

Then $X_{j}$ is (approximately) distributed as a uniform over $\{0, \ldots, m-1\}$

$\Rightarrow$ Define $U_j := X_j / m$, $U_j \approx \text{Uniform}[0, 1]$ 

```{r, include=TRUE}
# 'yield' equivalence: a closure preserving the generator's internal state
lcg <- function(modulus, a, c, seed) {
  state <- seed
  function() {
    state <<- (a * state + c) %% modulus
    return(state)
  }
}
```

```{r, include=TRUE}
mod <- 128
mult <- 15
increment <- 25
seed <- 1

samples1 <- numeric(10000)
gen <- function(x) {
  lcg(mod, mult, increment, x)
}

g <- gen(seed)
for (i in 1:10000) {
  temp <- g()
  samples1[i] <- temp / mod
}

# Plot histogram using R Base
hist(
  samples1, 
  breaks = 100, 
  col = "steelblue", 
  main = "", 
  xlab = "Value", 
  ylab = "Frequency"
)
```

```{r, include=TRUE}
mod <- 2^31 
mult <- 1103515245
increment <- 12345
seed <- 123

samples2 <- numeric(10000)

g <- gen(seed)
for (i in 1:10000) {
  temp <- g()
  samples2[i] <- temp / mod
}

# Plot histogram using R Base (Scott's rule for binning)
hist(
  samples2, 
  breaks = "Scott", 
  col = "steelblue", 
  main = "", 
  xlab = "Value", 
  ylab = "Frequency"
)
```

# Compare two PRNGs using the Kolmogorov - Smirnov test

Given samples $x_1, \ldots, x_n$ let  $F_n(t) = \frac{1}{n} \sum_{i=1}^n \mathbb I[x_i < t]$ the empirical CDF.

The Kolmogorov-Smirnov test compares the emprical CDF with the expected one, i.e. tests

$$
 H_0: \lim_{n\rightarrow\infty} F_n = F_0 \qquad \text{vs. } \qquad H_1:  \lim_{n\rightarrow\infty} F_n \neq F_0
$$

Is based on the Kolmogorov-Smirnov distance

$$
 KS(F_n, F_0) = \sup_x |F_n(x) - F_0(x)|
$$
```{r, include=TRUE}
# Kolmogorov-Smirnov test against uniform distribution U(0, 1)
ks.test(samples1, "punif")
```

```{r, include=TRUE}
# Kolmogorov-Smirnov test against uniform distribution U(0, 1)
ks.test(samples2, "punif")
```

```{r, include=TRUE}
# Kolmogorov-Smirnov test against uniform distribution U(0, 1)
ks.test(runif(1000), "punif")
```

# The Inverse CDF method

To generate a random value from $X$ we can:

1. Simulate $U \sim \text{Uniform}[0, 1]$
2. Return $F_X^{-1}(U)$

N.B. this works also for non continuous random variables!!

## Example 1 - Exponential distribution

$$ X \sim \text{Exp}(\lambda) \quad\Rightarrow\quad F_X(x) = 1 - \exp(-\lambda x) $$

```{r, include=TRUE}
exponential_rng <- function(lambda, size = 1){
  u <- runif(size)
  -log(1 - u)/lambda
}

hist(exponential_rng(2,1000), breaks = "Scott", 
  col = "steelblue", , probability=TRUE, xlim = c(0,4), ylim = c(0,1.5), main = '', xlab = '', ylab = '')
curve(dexp(x, 2), add = T, col = 'darkorange', lwd = 2)
```

Observe that

$$
1 - U \sim \text{Uniform}[0, 1]
$$

so we can speed up the code writing

```{r, include=TRUE}
exponential_rng <- function(lambda, size = 1){
  u <- runif(size)
  -log(u)/lambda
}

hist(exponential_rng(2,1000), breaks = "Scott", 
  col = "steelblue", probability=TRUE, xlim = c(0,4), ylim = c(0,1.5), main = '', xlab = '', ylab = '')
curve(dexp(x, 2), add = T, col = 'darkorange', lwd = 2)
```

## Example 2 - the truncated Normal distribution

We say that $X$ follows a truncated normal distribution with parameters $\mu, \sigma^2, a, b$ if its probability density function is

$$
    f_X(x) = \frac{1}{Z(\mu, \sigma^2, a, b)}\exp\left( -\frac{(x - \mu)^2}{2 \sigma^2} \right) \mathbb I[x \in (a, b)] \propto \exp\left( -\frac{(x - \mu)^2}{2 \sigma^2} \right) \mathbb I[x \in (a, b)]
$$
<!--
We can also compute the normalizing constant:

$$
    Z(\mu, \sigma^2, a, b) = P(Y \in [a, b])
$$
where $Y \sim \mathcal{N} (\mu, \sigma^2)$

Hence 
$$
    Z(\mu, \sigma^2, a, b) = \Phi\left( \frac{b - \mu}{\sigma} \right)  - \Phi\left( \frac{a - \mu}{\sigma} \right)
$$

The cumulative distribution function is then

$$
        F_X(x) =  \begin{cases}
        0 & \text{if } x < a \\
        \frac{1}{Z(\mu, \sigma^2, a, b)} \times \left( \Phi\left( \frac{x - \mu}{\sigma} \right)  - \Phi\left( \frac{a - \mu}{\sigma} \right) \right) & \text{if } a \geq x \geq b \\
        1 & \text{if } x > b
        \end{cases}
$$


The quantile function is then

$$
 u = \frac{1}{Z(\mu, \sigma^2, a, b)} \times \left( \Phi\left( \frac{x - \mu}{\sigma} \right)  - \Phi\left( \frac{a - \mu}{\sigma} \right) \right)
$$

$$
 Z(\mu, \sigma^2, a, b) u + \Phi\left( \frac{a - \mu}{\sigma} \right)  =  \Phi\left( \frac{x - \mu}{\sigma} \right) 
$$

$$
 \Phi^{-1} \left( Z(\mu, \sigma^2, a, b) u + \Phi\left( \frac{a - \mu}{\sigma} \right) \right)   \sigma + \mu  =  x
$$ -->

```{r, include=TRUE}
library(ggplot2)

# Truncated normal sampler via inverse CDF method
trunc_normal_rng <- function(mu, sigma, a, b, size) {
  u <- runif(size)
  z <- pnorm((b - mu) / sigma) - pnorm((a - mu) / sigma)
  qnorm(z * u + pnorm((a - mu) / sigma)) * sigma + mu
}

# Grid of points and theoretical PDF calculation
xx <- seq(-3, 10, length.out = 1000)
pdf_vals <- dnorm(xx, mean = 0, sd = 2) / (pnorm(10 / 2) - pnorm(-3 / 2))

# Generate sample data and create data frames for ggplot2
samples <- trunc_normal_rng(mu = 0, sigma = 2, a = -3, b = 10, size = 10000)
df_samples <- data.frame(x = samples)
df_pdf <- data.frame(x = xx, y = pdf_vals)

# Compute binwidth using Scott's rule
bw <- 3.5 * sd(samples) / (length(samples)^(1/3))

# Plot histogram with overlaid theoretical density
ggplot() +
  geom_histogram(
    data = df_samples, 
    aes(x = x, y = after_stat(density)), 
    binwidth = bw, 
    fill = "steelblue", 
    color = "black", 
    alpha = 0.5
  ) +
  geom_line(data = df_pdf, aes(x = x, y = y), color = "darkorange", linewidth = 1.2) +
  theme_minimal() +
  labs(x = "x", y = "Density")
```
